---
title: "Rasters"
output:
  html_notebook: 
    css: https://ucanr-igis.github.io/caladaptr-res/assets/nb_css01.css
    includes:
      after_body: https://ucanr-igis.github.io/caladaptr-res/assets/nb_footer01.html
---

# Overview

This notebook will demonstrate how to download Cal-Adapt data as rasters.

# Setup

Load caladaptR and the other package we're going to need. (If you haven't installed these yet, see this [setup script](https://github.com/ucanr-igis/caladaptr-res/blob/main/docs/workshops/ca_intro_apr22/scripts/caladaptr_setup.R)). 

```{r chunk01, message=FALSE, warning=FALSE, results='hold'}
library(caladaptr)
library(units)
library(dplyr)
library(lubridate)
library(sf)
library(stars)
library(conflicted)
conflict_prefer("filter", "dplyr", quiet = TRUE)
conflict_prefer("count", "dplyr", quiet = TRUE)
conflict_prefer("select", "dplyr", quiet = TRUE)
```

\

# Downloading Rasters

The same API Request object can be used to get raster data if you feed it into `ca_getrst_stars()`.

For additional info on downloading and analyzing rasters, see the 3 articles on [Downloading Rasters](https://ucanr-igis.github.io/caladaptr/articles/rasters-pt1.html).

Below we get a raster of observed historic temperature data for the Sierra climate region:

```{r chunk02, cache = FALSE}
sierra_cap <- ca_loc_aoipreset(type = "climregions", idfld = "name", idval = "Sierra") %>% 
  ca_livneh(TRUE) %>% 
  ca_period("year") %>% 
  ca_cvar("pr") %>% 
  ca_years(start = 1970, end = 2010)

sierra_cap

sierra_cap %>% ca_preflight(check_for = "getrst")

plot(sierra_cap, locagrid = TRUE)
```

\

To fetch the data as TIFs, use :

```{r chunk03}
tiff_dir <- "./data"

sierra_tiff_fn <- sierra_cap %>% 
  ca_getrst_stars(out_dir = tiff_dir, mask = TRUE, quiet = TRUE, overwrite = FALSE)
```

\

**Pro Tip:**

 - to avoid downloading the same TIF multiple times, use the same output directory and set `overwrite = FALSE`
 
\
 
`ca_getrst_stars()` returns a vector of TIF files that were downloaded. To work with them, you next have to load them back into R as stars objects (space-time arrays) using `ca_stars_read()`:

```{r chunk04, paged.print = FALSE}
sierra_stars_lst <- ca_stars_read(sierra_tiff_fn)
length(sierra_stars_lst)
sierra_stars_lst[[1]]
```

\

# Plot

To plot a stars objects, you have to decide which layer(s) to plot. In this case, each layer represents a year from 1970 to 2010. Below we plot 4 of the 40 years:

```{r chunk05, cache = FALSE}
plot(sierra_stars_lst[[1]] %>% slice(index = seq(1,40,length.out =4), along = "year"), 
     axes = TRUE,
     main = attributes(sierra_stars_lst[[1]])$ca_metadata$slug)
```

Not sure what the units are? You can double-check by viewing the metadata for the slug from the catalog:

```{r chunk06}
ca_catalog_search("pr_year_livneh")
```

\

# Subset by Dimension

Subsetting stars objects by a dimension is the counterpart to doing an attribute query in GIS. You can use dplyr filter function as you would a tibble. For example, to get the precipitation for only the 1990s:

```{r chunk07}
sierra_stars_lst[[1]] %>% filter(year >= 1990, year < 2000)
```

To get the values of a dimension, you can use

```{r chunk08}
(x_range <- sierra_stars_lst[[1]] %>% st_get_dimension_values("x"))
(yr_range <- sierra_stars_lst[[1]] %>% st_get_dimension_values("year"))
```

\

You can also use square bracket notation as you would an array. Just put an extra comma at the beginning of the bracket.

```{r chunk09}
sierra_stars_lst[[1]][ , , , which(yr_range >= 1990 & yr_range < 2000)]
```

\

# Spatial Selection

To spatially select, you can put a SF object in the first slot of the square bracket notation. Let's get the precipitation history for Yosemite NP:

```{r chunk10}
(ynp_bnd_sf <- st_read("https://raw.githubusercontent.com/ucanr-igis/caladaptr-res/main/geoms/ynp_bnd.geojson"))
plot(ynp_bnd_sf$geometry, axes = TRUE)
```

\

To do a spatial selection, put the sf object in the first slot in the square brackets:

```{r chunk11}
ynp_pr_stars <- sierra_stars_lst[[1]][ynp_bnd_sf , , , ]
ynp_pr_stars
```

\

Plot to verify:

```{r chunk12}
{plot(ynp_pr_stars %>% slice(index = 1, along = "year"), 
     axes = TRUE, main = "Average Daily Precipitation in YNP (mm), 1970", reset = FALSE)
plot(ynp_bnd_sf$geometry, border = "red", lwd = 2, add = TRUE)}
```

\

# Convert Values / Raster Algebra

Math operators (like `* / - +`) do 'pixelwise' operations on stars objects, just like regular arrays.

To compute the yearly summary in inches, rather than daily average in mm, we just multiply the stars object by a conversion constant. A mm is 0.0393701 inches, so to go from daily averages in mm to annual total in inches:

```{r chunk13}
(ynp_prtotal_stars <- ynp_pr_stars * 0.0393701 * 365)
```

\

Plot the first year to verify:

```{r chunk14}
{plot(ynp_prtotal_stars %>% slice(index = 1, along = "year"), 
     axes = TRUE, main = "Total Precipitation in YNP (in), 1970", reset = FALSE)
plot(ynp_bnd_sf$geometry, border = "red", lwd = 2, add = TRUE)}

```

\

# Aggregate Across Time and Space

To take the average precip per pixel for the entire time range, use `st_apply()`. The `MARGIN` argument should be the indices of the dimensions you want to *keep*:

```{r chunk15}
ynp_prtotal_stars %>% 
  st_apply(MARGIN = 1:2, FUN = mean) %>% 
  plot(axes = TRUE, main = "Average Total Annual Precipitation in YNP (in), 1970-2010")
```

\

To take the average total precip of all pixels in the park per year, we simply change the value of `MARGIN` to 3 (because year is the third dimension). We also have to add `na.rm = TRUE` to tell the `mean` function to ignore all the NA values outside the park boundary:

```{r chunk16}
(ynp_pr_annual_avg_stars <- ynp_prtotal_stars %>% 
  st_apply(MARGIN = 3, FUN = mean, na.rm = TRUE))
```

\

stars objects can be converted to data frames:

```{r chunk17}
ynp_pr_annual_avg_df <- ynp_pr_annual_avg_stars %>% as.data.frame
plot(ynp_pr_annual_avg_df, type="b", main = "YNP Total Precip Averaged Across Park (in)")
```

\

To plot the distribution of values, we can grab the individual values with `[]` (just like an array) and make a histogram:

```{r chunk18}
hist(ynp_prtotal_stars, main = "Distribution of Total Annual Precip in YNP")
```


There is a **lot** more you can do with rasters, including pixel summaries, combining them into higher dimensional data cubes, spatially mosaicing them, etc. For more info, see the Rasters articles on the [website](https://ucanr-igis.github.io/caladaptr/).
 
\

#### Your Turn

Download historic precipitation data for the county where you live or work. [[Answer](https://bit.ly/3m7BcWM)]
 
```{r chunk19, paged.print = FALSE, cache=FALSE}
## Your answer here

```
 

