---
title: "Notebook 2: Data Munging Techniques"
output: html_notebook
---

# Introduction

In this exercise we'll explore more data munging techniques for climate data, focusing on daily data. 

# Setup

First we load the packages we need for this exercise, which include `caladaptr`, `dplyr`, and `sf`:

```{r message = FALSE}
library(caladaptr)
library(sf)
library(dplyr)
library(lubridate)
library(units)
library(ggplot2)
library(tidyr)
```

`dplyr` in particular has a number of very generic function names, so we tell R which one we want it to use using the conflicted package:

```{r}
library(conflicted)
conflict_prefer("filter", "dplyr", quiet = TRUE)
conflict_prefer("count", "dplyr", quiet = TRUE)
conflict_prefer("select", "dplyr", quiet = TRUE)
```

TIP: We can also customize the color of the output of caladaptr functions:

```{r}
options(ca_accent1 = crayon::yellow$bold)
options(ca_accent2 = crayon::yellow)
options(ca_accent3 = crayon::cyan$bold)
options(ca_accent4 = crayon::cyan)
options(ca_message = crayon::yellow)
options(ca_success = crayon::green)
```

# Grab 10 years of daily data for a point

In this example, we'll explore daily projected climate data for a point location. We select the UC [**Lindcove Research and Extension Center**](http://lrec.ucanr.edu/) (LREC), a field station in the San Joaquin Valley which has been a leading center for citrus research for decades. The trees are getting old and need to be replaced soon so its worth asking - is citrus production still going to viable in this part of California mid-century? 

## Create the API Request

Let's start by getting 20 years worth of daily data for the 4 priority GCMs and 2 RCPs. 

```{r}
lrec_prj_cap <- ca_loc_pt(coords = c(-119.060, 36.359), id = 1) %>% 
  ca_period("day") %>% 
  ca_gcm(gcms[1:4]) %>% 
  ca_scenario(c("rcp45", "rcp85")) %>% 
  ca_cvar(c("tasmin", "tasmax")) %>% 
  ca_years(start = 2040, end = 2060)
  
lrec_prj_cap
```

```{r}
plot(lrec_prj_cap)
```

## Fetch Data

Now we're ready to fetch data:

```{r}
lrec_prj_tbl <- lrec_prj_cap %>% ca_getvals_tbl()

## backup: lrec_prj_tbl <- readRDS("data/lrec_prj_tbl.rds")

dim(lrec_prj_tbl)
head(lrec_prj_tbl)
```

# Create a Box Plot of Temperature Values by Month

First add columns for Fahrenheit and month.

```{r}
lrec_prj_tempf_tbl <- lrec_prj_tbl %>% 
  mutate(temp_f = set_units(val, degF), month = month(dt), year = year(dt))
head(lrec_prj_tempf_tbl)
```

For each month, let make a box plot of the temperature values for each emission scenario, treating all GCMs as equally likely:

```{r}
ggplot(lrec_prj_tempf_tbl, aes(x = as.factor(month), y = as.numeric(temp_f))) + 
  geom_boxplot() +
  facet_grid(scenario ~ .) +
  labs(title = "Daily Temperature Range by Month", x = "month", y = "temp (F)",
       subtitle = "Lindcove REC, 4 GCMs combined, 2040-2060")
```
## YOUR TURN

Create a histogram for each RCP of maximum temperature values in July.

Hint: the geom function for a histogram is `geom_histogram()`. [Solution](https://bit.ly/2JPsXzG): 

```{r}

```

# Count extreme heat days

An extreme heat day is generally identified when the maximum temperature exceeds a threshold. The threshold can be chosen based on historical patterns, or a biophysical process. For this example, we'll select 105.

Let's count the total number of days the temperature exceeded 105. We start by adding a logical column whether the temperature exceeded our threshold. We'll also throw away the minimum temperature values. 

```{r}
lrec_prj_hot_tbl <- lrec_prj_tempf_tbl %>% 
  filter(cvar == "tasmax") %>% 
  mutate(really_hot = (temp_f >= set_units(105, degF))) %>% 
  select(-spag, -val, -month)

head(lrec_prj_hot_tbl)
```

We can count the number of extreme heat days with a simple expression:

```{r}
num_hot_days <- lrec_prj_hot_tbl %>%
  group_by(scenario, really_hot) %>% 
  count()

num_hot_days
```

We can improve the readability of this table by making each scenario a separate column. This is an example of [pivoting](https://tidyr.tidyverse.org/articles/pivot.html), which you can handle using `tidyr::pivot_wider`.

```{r}
num_hot_days %>% pivot_wider(names_from = scenario, values_from = n)
```

## YOUR TURN

Count the number of extreme heat days using a threshold of 110 Fahrenheit. Compute the mean number of extreme heat days per month.

```{r}

```




# Count Consecutive Extreme Heat Days

Sometimes what matters most is not just the number of extreme heat days, but the number of consecutive extreme heat days lasting *n* or more days (i.e., an extreme heat spell). We can identify consecutive extreme heat days by using `rle` (run-length encoding), which simply finds groups of repeating values in a vector (as a primitive form of compression). But we have to chop up our data into 'runs' (year + gcm + scenario):

```{r}
## Group the rows by temperature values by year, GCM, and scenario
temp_by_gcm_year_scenario <- lrec_prj_hot_tbl %>% 
  select(year, gcm, scenario, dt, really_hot) %>% 
  group_by(year, gcm, scenario)

## Split each group of rows into its own tibble
temp_by_gcm_lst <- temp_by_gcm_year_scenario %>% group_split()

## For each tibble, run the really_hot column through rle()
rle_lst <- lapply(temp_by_gcm_lst, function(x) x %>% pull(really_hot) %>% rle())

## Define the minimum heat spell length
spell_length <- 3

## Count up the number of runs of really_hot=TRUE lasting >= spell_length days  
num_spells_vec <- sapply(rle_lst, function(x) sum( x$lengths[x$values] >= spell_length))

## Add the number of spells column to the grouping keys (year, gcm, scenario)
num_spells_tbl <- group_keys(temp_by_gcm_year_scenario) %>% 
  mutate(num_heat_spells = num_spells_vec)

## View the avg number of heat spells per scenario
num_spells_tbl %>% 
  group_by(scenario) %>% 
  summarise(mean_heat_spells_per_year = mean(num_heat_spells))
```

# Display a Trend Line and Ensemble Range

In this part we'll see how to add trend lines and the range of values of the 32-GCM ensemble to a time series plot.

First, we fetch 20 years of annual data for LREC, 4 GCMs, and 2 scenarios. Note below how we can combine the API reqest, data fetching, and data munging in one expression.

```{r}
lrec_prj_year_tbl <- 
  ca_loc_pt(coords = c(-119.060, 36.359), id = 1) %>% 
  ca_period("year") %>% 
  ca_gcm(gcms[1:4]) %>% 
  ca_scenario(c("rcp45", "rcp85")) %>% 
  ca_cvar(c("tasmin", "tasmax")) %>% 
  ca_years(start = 2040, end = 2060)%>% 
  ca_getvals_tbl() %>% 
  mutate(temp_f = set_units(val, degF), year = year(dt)) %>% 
  group_by(year, scenario) %>% 
  summarise(mean_temp = mean(temp_f))

## backup: lrec_prj_year_tbl <- readRDS("data/lrec_prj_year_tbl.rds")

dim(lrec_prj_year_tbl)
head(lrec_prj_year_tbl)
```

ggplot can create various types of trend lines by simply adding `geom_smooth()` to your plot definition: 

```{r}
ggplot(data = lrec_prj_year_tbl, 
       aes(x = year, y = as.numeric(mean_temp))) +
  geom_line() +
  geom_smooth(method = "lm") +
  facet_grid(scenario ~ .) +
  labs(title = "LREC Mean Annual Temperature", x = "year", y = "temp (F)",
       subtitle = "2040-2060, 4 GCMs averaged")

```

The Cal-Adapt API also has climate data generated from an "ensemble GCM", which includes all 32 GCMs in the CMIP5 project. We can show this in the background to get a feel for how close or how far off the 4GCMs we selected are from the main models.

First we grab the ensemble data:

```{r}
lrec_ens_year_tbl <- 
  ca_loc_pt(coords = c(-119.060, 36.359), id = 1) %>% 
  ca_period("year") %>% 
  ca_gcm("ens32max") %>% 
  ca_scenario(c("rcp45", "rcp85")) %>% 
  ca_cvar(c("tasmin", "tasmax")) %>% 
  ca_years(start = 2040, end = 2060)%>% 
  ca_getvals_tbl() %>% 
  mutate(temp_f = set_units(val, degF), year = year(dt)) %>% 
  select(cvar, gcm, scenario, year, temp_f)

## backup: lrec_ens_year_tbl <- readRDS("data/lrec_ens_year_tbl.rds")

lrec_ens_year_tbl
```

To display the min and max as a shaded background layer, we have to split them into different columns using `pivot_wider()`: 

```{r}
lrec_ens_min_max <- lrec_ens_year_tbl %>%
  pivot_wider(names_from = cvar, 
              values_from = temp_f,
              id_cols = c(year, scenario))

head(lrec_ens_min_max)
```

Now we can plot them as a `geom_ribbon()`:

```{r}
ggplot() +
  geom_ribbon(data = lrec_ens_min_max, 
              mapping = aes(x = year, ymin = as.numeric(tasmin), ymax = as.numeric(tasmax)), 
              fill = "gray70") +
  geom_line(data = lrec_prj_year_tbl,
            mapping = aes(x = year, y = as.numeric(mean_temp))) +
  facet_grid(scenario ~ .) +
  labs(title = "LREC Annual Mean Temperature", x = "year", y = "temp (F)",
       subtitle = "2040-2060",
       caption = "Average of 4 GCMs with 32-GCM ensemble range in background")

```
