---
title: "Importing Windspeed Datasets"
date: "July 17, 2022"
output:
  html_notebook: 
    css: https://ucanr-igis.github.io/caladaptr-res/assets/nb_css01.css
    includes:
      before_body: https://ucanr-igis.github.io/caladaptr-res/assets/nb-hdr_cde-rcps.html
---

\

This Code Recipe Notebook demonstrates how to import modeled windspeed datasets from Cal-Adapt into R using [caladaptR](https://ucanr-igis.github.io/caladaptr/).

# Load packages

```{r setup, message = FALSE}
library(dplyr, quietly = TRUE)
library(sf)
library(stars)
library(caladaptr)
packageVersion("caladaptr")
```

# Look for wind raster series 

The following command doesn't work in a Notebook, but if you run it at the console it will open up the raster series catalog in RStudio pane. You can then use the 'Filter' button to search for datasets which have the word 'wind' in the name field.

```{r}
# ca_catalog_rs() %>% View()
```

\


```{r}
windrs_df <- ca_catalog_search("wind")
```

\

View just the slugs:

```{r}
windrs_df$slug
```

\

It would appear there are raster series datasets for modelled windspeed for the first four 'priority' GCM models (2006-2100), as well as modeled historical data (1950-2005) (which is what you should use to compare future modeled windspeed). There is also a raster series for daily maximum wind speed historical observed (`wspeed_day_hadisd`).

\

# Example 1: Query Modeled Future Windspeed Data

Create an API request for future modeled wind speed. Because there isn't a convenience function to construct an API request object for windspeed, we can use the slugs:

```{r}
yose_wind_cap <- ca_loc_pt(coords = c(-119.591, 37.745)) %>% 
  ca_slug(c("wspeed_day_CanESM2_rcp45",
            "wspeed_day_CanESM2_rcp85",
            "wspeed_day_CNRM-CM5_rcp45",
            "wspeed_day_CNRM-CM5_rcp85",
            "wspeed_day_HadGEM2-ES_rcp45",
            "wspeed_day_HadGEM2-ES_rcp85",
            "wspeed_day_MIROC5_rcp45",
            "wspeed_day_MIROC5_rcp85")) %>% 
  ca_years(start = 2070, end = 2099)

yose_wind_cap
```

\

Pre-flight checks:

```{r}
yose_wind_cap %>% ca_preflight()
yose_wind_cap %>% plot()
```

\

Grab values:

```{r}
yose_wind_tbl <- yose_wind_cap %>% 
  ca_getvals_tbl(quiet = TRUE)
```

\

Inspect the results:

```{r}
glimpse(yose_wind_tbl)
```

\

# Example 2: Download Modeled Windspeed Data for an Entire County as a TIF

Next, we grab historic observed wind speed for Mono County.

1) Find thing is we need to find the fips code for Mono County:

```{r}
mono_cnty_fips <- ca_aoipreset_geom("counties", quiet = TRUE) %>% 
  st_drop_geometry() %>% 
  filter(grepl("Mono", name)) %>% 
  pull(fips)

mono_cnty_fips
```

2) Create the API request:

```{r}
mono_wspeed_cap <- ca_loc_aoipreset(type = "counties",
                                        idfld = "fips",
                                        idval = "06051") %>% 
  ca_slug(c("wspeed_day_CanESM2_rcp45",
            "wspeed_day_CNRM-CM5_rcp45",
            "wspeed_day_HadGEM2-ES_rcp45",
            "wspeed_day_MIROC5_rcp45")) %>% 
  ca_years(start = 2070, end = 2099)

mono_wspeed_cap

```

\

Preflight checks:

```{r}
mono_wspeed_cap %>% ca_preflight()
mono_wspeed_cap %>% plot(locagrid = TRUE)
```

\

Get the data as TIFs:

```{r}
tiff_dir <- tools::R_user_dir("caladaptr", which = "data") %>% file.path("Mono")
if (!file.exists(tiff_dir)) dir.create(tiff_dir, recursive = TRUE)

mono_wspeed_fn <- mono_wspeed_cap %>% 
  ca_getrst_stars(out_dir = tiff_dir)

mono_wspeed_fn
```

\

Read the windspeed data back in:

```{r}
mono_wspeed_lst <- ca_stars_read(mono_wspeed_fn)
names(mono_wspeed_lst)

mono_wspeed_lst[[1]]
```

\

Plot wind speed for the first week of January for the first raster: `wspeed_day_CanESM2_rcp45_fips-06051`:

```{r}
plot(mono_wspeed_lst[[1]] %>% slice(index = 1:7, along = "date"), 
     axes = TRUE,
     main = st_get_dimension_values(mono_wspeed_lst[[1]], "date")[1:7])
```



